---
title: "Reproducible Research: Peer Assessment 1"
output:
    html_document:
        keep_md: true
---

## Loading and preprocessing the data

First we need to make sure we reference all required libraries
```{r references,results='hide'}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
```

Now we need to load the data and inspect it.
```{r getdata}
activity <- read.csv("activity.csv")
str(activity)
summary(activity)
```

The 'date' variable was read in as a character Factor and should be a Date datatype.
```{r preprocessdates}
activity$date <- as.POSIXct(activity$date)
str(activity)
```


##Steps per Day Analysis

Our first analysis is around the total # of steps per day. Let's create a new dataset to hold these values.
```{r createdailysteps}
dailySteps <- activity %>%
    filter(!is.na(steps)) %>%
    group_by(date) %>%
    summarize(steps=sum(steps))
str(dailySteps)
```

Let's see a histogram of the # of steps per day
```{r dailystepshisto,fig.height=6,fig.width=8}
ggplot(dailySteps, aes(x=steps)) +
    geom_histogram(binwidth=1000, color="black") +
    labs(title="Steps per Day", 
         x="# of Steps", 
         y="# of Days"
         )
```

Let's look at the mean and median # of steps:
```{r dailystepsmean,echo=FALSE,results='hide'}
dlyStepsMean <- mean(dailySteps$steps)
dlyStepsMedian <- median(dailySteps$steps)
```
* Mean # of steps: `r format(dlyStepsMean, scientific=100, big.mark = ",")`
* Median # of steps: `r format(dlyStepsMedian, big.mark = ",")`

## Average Daily Activity Pattern Analysis

We want to look at what the average daily activity pattern looks like.  First, lets get a dataset which shows the total and average # of steps for each 5-minute interval
```{r avgpattern}
intervalSteps <- activity %>%
    filter(!is.na(steps)) %>%
    group_by(interval) %>%
    summarize(totSteps=sum(steps), avgSteps=mean(steps))
str(intervalSteps)
```

Let's chart a time-series graph to see how the daily pattern.
```{r avgpatternchart,fig.height=6,fig.width=8}
ggplot(intervalSteps, aes(x=interval,y=avgSteps,group=1)) +
    geom_line(color="blue") +
    labs(title="Average Steps by Interval",
         xlab="5-min Interval",
         ylab="Avg # of Steps"
         )
```

