---
title: "Reproducible Research: Peer Assessment 1"
output:
    html_document:
        keep_md: true
---

## Loading and preprocessing the data

First we need to make sure we reference all required libraries
```{r references,results='hide'}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
```

Now we need to load the data and inspect it.
```{r getdata}
activity <- read.csv("activity.csv")
str(activity)
summary(activity)
```

The 'date' variable was read in as a character Factor and should be a Date datatype.
```{r preprocessdates}
activity$date <- as.POSIXct(activity$date)
str(activity)
```


##Steps per Day Analysis

Our first analysis is around the total # of steps per day. Let's create a new dataset to hold these values.
```{r createdailysteps}
dailySteps <- activity %>%
    filter(!is.na(steps)) %>%
    group_by(date) %>%
    summarize(steps=sum(steps))
str(dailySteps)
```

Let's see a histogram of the # of steps per day
```{r dailystepshisto,fig.height=6,fig.width=8}
ggplot(dailySteps, aes(x=steps)) +
    geom_histogram(binwidth=1000, color="black") +
    labs(title="Steps per Day", 
         x="# of Steps", 
         y="# of Days"
         )
```

Let's look at the mean and median # of steps:
```{r dailystepsmean,echo=FALSE,results='hide'}
dlyStepsMean <- mean(dailySteps$steps)
dlyStepsMedian <- median(dailySteps$steps)
```
* Mean # of steps: `r format(dlyStepsMean, scientific=100, big.mark = ",")`
* Median # of steps: `r format(dlyStepsMedian, big.mark = ",")`

## Average Daily Activity Pattern Analysis

We want to look at what the average daily activity pattern looks like.  First, lets get a dataset which shows the total and average # of steps for each 5-minute interval
```{r avgpattern}
intervalSteps <- activity %>%
    filter(!is.na(steps)) %>%
    group_by(interval) %>%
    summarize(totSteps=sum(steps), avgSteps=mean(steps))
str(intervalSteps)
```

```{r computehighavgperiod,echo=FALSE,results='hide'}
highestPeriod <- intervalSteps[intervalSteps$avgSteps==max(intervalSteps$avgSteps),]$interval
```


Let's chart a time-series graph to see how the daily pattern.
```{r avgpatternchart,fig.height=6,fig.width=8}
ggplot(intervalSteps, aes(x=interval,y=avgSteps,group=1)) +
    geom_line(color="blue") +
    geom_vline(xintercept = highestPeriod, color="green") +
    labs(title="Average Steps by Interval",
         xlab="5-min Interval",
         ylab="Avg # of Steps"
         )
```

The 5-minute period which has the maximum average # of steps across all dates is period **`r highestPeriod`** marked by the green line.


## Imputing Missing Values

There are a number of days/intervals where there are missing values. The presence of missing days may introduce bias into some calculations or summaries of the data.

```{r nbrnas}
incompleteRows <- nrow(activity[!complete.cases(activity), ])
incompleteRows
```

So, we know there are **`r format(incompleteRows, big.mark=",")`** incomplete rows in the dataset.

We need to replace all missing values (NAs) with a value to help avoid bias.  We'll use the average of the # of steps for the 5-minute period the missing value is associated with.
```{r imputemeans}
impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
fixedActivity <- activity %>%
    group_by(interval) %>%
    mutate(stepsNew = impute.mean(steps)) %>%
    select(-steps) %>%
    inner_join(activity, by=c('interval','date')) %>%
    select(-steps) %>%
    rename(steps = stepsNew)
summary(fixedActivity)
incompleteRows <- nrow(fixedActivity[!complete.cases(fixedActivity), ])
incompleteRows
```

No we see all rows are complete!

Let's make another Histogram now to see how the daily steps look.
```{r imputedhistogram,fig.height=6,fig.width=8}
fixedDailySteps <- fixedActivity %>%
    group_by(date) %>%
    summarize(steps=sum(steps))
ggplot(fixedDailySteps, aes(x=steps)) +
    geom_histogram(binwidth=1000, color="black") +
    labs(title="Steps per Day", 
         x="# of Steps", 
         y="# of Days"
         )
```

Let's look at the mean and median # of steps after the fix is applied compared to the original values:
```{r fixeddailystepsmean,echo=FALSE,results='hide'}
dlyFixedStepsMean <- mean(fixedDailySteps$steps)
dlyFixedStepsMedian <- median(fixedDailySteps$steps)
```
* Mean # of steps: Original = `r format(dlyStepsMean, scientific=100, big.mark = ",")`; New = `r format(dlyFixedStepsMean, scientific=100, big.mark = ",")`
* Median # of steps: Original = `r format(dlyStepsMedian, big.mark = ",")`; New = `r format(dlyFixedStepsMedian, big.mark = ",")`

So, it appears that imputing the means had little affect on the estimates of total daily steps!

## Activity Patterns by Weekday

Now let's look at how the activity patterns are affected by whether it's a weekday or a weekend.  First we need to add a factor indicating if it is a weekday or weekend.
```{r addweekdaysgroup}
# use lubridate package to help identify the weekday of the date
suppressPackageStartupMessages(library(lubridate))
fixedActivity <- fixedActivity %>% mutate(weekdayGroup=as.factor(ifelse(wday(date) %in% c(1,7),"weekend","weekday")))
summary(fixedActivity$weekdayGroup)
```


Let's compare the average # of steps per 5-minute interval across the weekday groups.
```{r chartweekdaysteps}
weekdayIntervalSteps <- fixedActivity %>%
    group_by(interval, weekdayGroup) %>%
    summarize(stepsAvg=mean(steps),steps=sum(steps))
ggplot(weekdayIntervalSteps, aes(x=interval,y=stepsAvg,color=weekdayGroup)) +
    facet_wrap("weekdayGroup", ncol=1) +
    geom_line()
```


